 Quantile Regression Analysis for Statin Effects on Body mass Index

### 2021-02-12 manuscript/manuscript.Rmd
updating files
update

---
title: " Quantile Regression Analysis for Statin effects on glucose level"
output: 
    bookdown::pdf_document2:  
        keep_tex: true
bibliography: cite.bib
link_citations: true
geometry: margin=1in
fontsize: 12pt
fig_caption: yes
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
---
\section{introduction}

---
# here's a comment
# ```{r}
# Body Mass Index (BMI) plays an important rule in predicting heart disease #risk[@katbody]. Our goal is to compare 90Th quantile regression for BMI in Diabetes and #prediabetes population. Quantile regression has many succesfyl application in ecology #where different factores interect in a complicated way that produce different variation #of one factor for different levels of another variabels[@cade2003].
# ```
---
Approximately, 18 million deaths annually per year are caused by cardiovascular diseases(CVD) and similar to this number of nonfatal cardiovascular events [@hay2017gbd].
For 2011, annual costs for CVD and stroke were $\$320.1$ billion, which is more than cancer cost. This cost include \$195.6  billion in direct costs (healthcare costs), and cost of future productivity loss is  \$124.5 [@mozaffarian2015executive].
Abnormal lipid ratio  is a risk  factor for CVD [@yusuf2004effect]. Lowering low-density lipoprotein (LDL) using statin reduces risk of cardiovascular diseases even with population with no CVD  [@yusuf2016]. 
It is known that diabetes is a risk factor for CVD, however, it is shown that elevated  glucose, prediabetes,is also a risk factor for CVD [@balkau2004prediction].

Statin use associated with high risk of a new onset diabetes in normoglcemic (HR 1.19,95% CI , 1.05 to 1.35) and in prediabetes (HR 1.24 95% CI , 1.11 to 1.38). On the other hand, overall mortality risks went down  in both normoglycemic (HR 0.70;
95 % CI, 0.66 to 0.80) and IFG patients (HR
0.77, 95 % CI, 0.64 to 0.91) with statin use [@castro2016]. For population with one or more risk factor for diabetes, statin use associate with 39% reduction in the primary end points (the hazard ratio (HR) 0.61), and a 28% (HR 1.28)increase in diabetes[@ridker2012].Pandya at el.[pandya2015] suggested the dis-utility associated with taking a pill daily, statin price, and the risk of statin-induced diabetes negatively impacted the cost effectiveness of statin use recommendation.

@@

$E\rho_{\tau}(X-\hat{x})=(\tau-1)\int_{-\infty}^{\hat{x}} (x-\hat{x})dF(x)+\tau\int_{\hat{x}}^{-\infty} (x-\hat{x})dF(x).$
Differentiating with respect to $\hat{x}$, we get
$$ 0  =(\tau-1)\int_{-\infty}^{\hat{x}} dF(x)+\tau\int_{\hat{x}}^{-\infty} dF(x)
=F(\hat{x})-\tau.$$ 
Due to monotnicity of the cumulative distribution function, any solution that satisfies $\{x:F(x)=\tau \}$ is a minimizing for the expected loss function.

@@

\section{Methods}
A multivariate quantile regression model is used to assess the characteristics of the   association variability in different quantiles of the conditional distribution of the glucose levels.
Moreover, the independent variables are Gender, Age, race, diabetes status, and statin use.

#Data
 The data used in this study is National Health and Nutrition Examination Survey data (NHANES)[@NHANES]. The survey examines a nationally representative sample of U.S. population.It focuses on  variety of health and nutrition measurements. In this study, we cumulated 6 cycles of NHANES data (2007–2018). There are 12,406 records. The dependent variables in our model are gender, race, age, BMI, total cholesterol, and statin use (yes or no). All type of statin are included. The included races are non-Hispanic white, non-Hispanic black, Hispanic, or other. BMI are classified into different categories according to underweight, 18.5 kg/m2; normal weight, 18.5 to 25 kg/m2; overweight, 25 to 30 kg/m2; obese, 30 to 35 kg/m2; and very obese more than 35 kg/m2.

| Syntax      | Male  | Femal     |
| :---        |    :----:   |          ---: |
| count |  5990      | 6416   |
| Mean of Age | 49.9       | 49.73   |
| glucose        |  112.618     |  106.47   |
| TC         | Title       | Here's this   |
| Statin use (ratio) | 0.198       | 0.181   |

@@

```{r, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "quant90.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "quant90.png"))
```
\section{Results}
The resulting estimate of effects on conditional mean of glucose level may not reflect the size and nature of theses effects on lower or upper quantile. For example, in the Figure \@ref(fig:resu1), the conditional mean shows the conditional mean effects of gender on glucose level is about 7. However, the disparity of the gender effects on lower tails is much smaller which is about 3 but the disparity is higher for the upper tail of the distribution somewhat larger than 10 mg/dL.

@@

```{r resu1, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig.png"))
```
 So,
 therefore
```{r, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig2.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig2.png"))
``` 
So,
 Age effect modeled as a quadratic factor. The age effect  is linear in general, see Figure \@ref(fig:resu3). The slope for the  left quantile is lower than the slope of the right quantile. This can be interpreted as age effect tends to increase over the entire range of conditional distribution of glucose in faster rate for the higher quantile than in the left quantile.In other word, diabetes population have higher rate of change with respect to age if compared to non diabetes population. It was found that there is  a significantly correlation between glucose and age[@tsaousis2014].

@@

```{r resu3, out.width = "80%", fig.align = "center", fig.cap = "Illustration of the quadratic age effect on glucose leveles for four different quantiles of the conditional glucose distribution.  The ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig3.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig3.png"))
```
Quadratic effect of total cholesterol on the conditional distribution of glucose is convex.  At the lower tail, decreasing glucose level by somewhat around 1.4 up to around 180 Cholesterol level. Then, started to increase glucose levels after TC beyond 180. At higher quantile, the reverse in the cholesterol effects is shifted to higher cholesterol levels which around 200, Figure \@ref(fig:resu4). The relation between total cholesterol and fasting glucose has been studied in [@tsaousis2014],[@chang2011]. They found that there is a positive correlation between the two groups on average.

@@

<!-- ```{r, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE} -->
<!-- library(knitr) -->
<!--  kable(dat[,2:3],caption="A nit table") -->
<!-- ``` -->

>>>>

---
title: " Quantile Regression Analysis for Statin Effects on Body mass Index"
output: 
    bookdown::pdf_document2:  
        keep_tex: true
bibliography: cite.bib
link_citations: true
geometry: margin=1in
fontsize: 12pt
fig_caption: yes
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
---

\section{introduction}

Body Mass Index (BMI) plays an important rule in predicting heart disease risk[@katbody]. Our Goal is to estimate the trend in the $\tau$th quantile of  BMI using set of independent predictors. BMI departs normality because of the skewnes on the right and it is shown by using Box-cox transformation [@flegal1999]. Estimation of underweight and overweight can be effected by the voilation of the normality. Quantile regression is a crucial tools used to estimate BMI with the dispersion of the association of predictors. We found that spline regressions produce different estimate than polynomial regressions.  Quantile regression has many successful application in ecology where different factors interact in a complicated way that produce different variation of one factor for different levels of another variables[@cade2003].

Approximately, 18 million deaths annually per year are caused by cardiovascular diseases(CVD) and similar to this number of nonfatal cardiovascular events [@hay2017gbd].
In 2011, annual costs for CVD and stroke were $\$320.1$ billion, which is more than cancer cost. This cost include \$195.6  billion in direct costs (healthcare costs), and cost of future productivity loss is  \$124.5 [@mozaffarian2015executive]. Important factors that associated with CVD are high body mass index (BMI) and abnormal lipid ratio  [@yusuf2004effect],[@anderson1991]. Lowering low-density lipoprotein (LDL) using statin or other drugs reduce risk of cardiovascular diseases even with population with no CVD  [@yusuf2016]. Ferrieres at el. [@ferrieres2018] studied effect of BMI on the choice of lipid-lowering treatment.It is been found that statin intensity increases with high level of BMI($\rho=0.13$).  
It is known that diabetes is a risk factor for CVD, however, it is shown that elevated  glucose, prediabetes,is also a risk factor for CVD [@balkau2004prediction].

Statin use  associates with high risk of elevating BMI level[@sugiyama2014]. It is been shown that statin users consume 192 additional calories per day which cause gaining a 6 lb-to-11 lb in a year. Statin users gain 1.3 units in the BMI measure while non-staining users gains 0.4 unit  .Moreover, consumption of fat in statin users raised by 14.4%.

Statin has negative impact on glucose level. The hazard ratio of a new onset diabetes in  normoglycemic statin users(HR 1.19,95% CI , 1.05 to 1.35) and in pre-diabetes (HR 1.24 95% CI , 1.11 to 1.38). On the other hand, overall mortality risks went down  in both normoglycemic (HR 0.70;
95 % CI, 0.66 to 0.80) and in prediabetes patients (HR
0.77, 95 % CI, 0.64 to 0.91) with statin use [@castro2016]. For population with one or more risk factor for diabetes, statin use associate with 39% reduction in the primary end points (the hazard ratio (HR) 0.61), and a 28% (HR 1.28)increase in diabetes[@ridker2012].Pandya at el.[@pandya2015] suggested the dis-utility associated with taking a pill daily, statin price, and the risk of statin-induced diabetes negatively impacted the cost effectiveness of statin use recommendation.

Several studies estimated the association of different factors like lipid-lowering drug  with the conditional mean of BMI using OLS or logistic regression model[@ferrieres2018]. The drawback in this study is not considering the dispersion of the association of covariates with conditional distribution of the dependent variables, BMI.

@@

$E\rho_{\tau}(X-\hat{x})=(\tau-1)\int_{-\infty}^{\hat{x}} (x-\hat{x})dF(x)+\tau\int_{\hat{x}}^{-\infty} (x-\hat{x})dF(x).$
Differentiating with respect to $\hat{x}$, we get
$$ 0  =(\tau-1)\int_{-\infty}^{\hat{x}} dF(x)+\tau\int_{\hat{x}}^{-\infty} dF(x)
=F(\hat{x})-\tau.$$ 
Due to monotonicity of the cumulative distribution function, any solution that satisfies $\{x:F(x)=\tau \}$ is a minimizing for the expected loss function.

@@

\subsection{Quantile Regression Technique }

\subsection{splines}
   A continuous predictor can be linear, say $ X$, or nonlinear, say $X^2$ depends on the relationship with the response variable. Most of the relations between the responses and predictors variables are complicated to the point that linear regressions are not suitable to model these relationships[@bruce2017]. For example, the response to different levels of drug doeses is not a linear relationship. Linear regressions can be generalized to deal with nonlinear effects. One approach is through including polynomial terms to regression equation. This approach is called  Polynomial regression. The mathematical model for $n$ degree polynomial regression is shown in the Eq\eqref{non}
 \begin{equation}\label{non}
 y=\beta_0+\beta_1 X+\beta_2 X^2+\cdots+\beta_nX^n.
 \end{equation}

 One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher order terms has negative impact on the model by introducing undesirable "wiggliness" in the regression equation.  Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsman's in ploting curves. Spline is a process of constructing a set of  piecewise continuous
polynomials that are smoothly connected at set of points in the range of predictor variable. The connection points are called knots i.e. splines are used to smoothly interpolate between certain points. Let $a,b$ ,and $c$ are the end points of the $x$-axis intervals. A smooth cubic spline function is defined by the following equation. 
\begin{equation}
f(X)=X\beta=\beta_{0}+\beta_{1}X+\beta_2X^2+\beta_{3}X^{3}+\beta_{4}(X-a)_{+}^3+\beta_{5}(X-b)_{+}^3+\beta_6(X-c)_{+}^3,
\end{equation}

where $$(u)_{+}=  
 \left\{
	\begin{array}{ll}
		u  & \mbox{if } u > 0 \\
		0 & \mbox{if } u \leq 0
	\end{array}
\right. $$

\section{Methods}
A multivariate quantile regression model is used to assess the characteristics of the   association variability in different quantiles of the conditional distribution of the body mass index.

 The independent variables in our model are gender, race, age, total cholesterol, Cholesterol drug uses (yes or no). All type of cholesterol drug are included including  statins. The included races are non-Hispanic white, non-Hispanic black, Hispanic, or other.

\section{Numerical Example}
\subsection{Data}
 The data used in this study is National Health and Nutrition Examination Survey data (NHANES)[@NHANES]. The survey examines a nationally representative sample of U.S. population.It focuses on  variety of health and nutrition measurements. The survey data are released every two years cycle.  In this study, we accumulated 6 cycles of NHANES data (2007–2018). We used 
 two data files: One contains demographic variables, such as age, sex, race, income, etc, and the other contains data that are related to body measurements, such as BMI, head circumference, etc. These files are marged by using the respondent sequence number (SEQN)
 There are  around 12,000 records. We selected  population age between 20 and 80.
 BMI are classified into different categories according to underweight, 18.5 kg/m2; normal weight, 18.5 to 25 kg/m2; overweight, 25 to 30 kg/m2; obese, 30 to 35 kg/m2; and very obese more than 35 kg/m2.

| Syntax      | Male  | Female     |
| :---        |    :----:   |          ---: |
| count |  5990      | 6416   |
| Mean of Age | 49.9       | 49.73   |
| BMI        |  28.549    |  29.379  |
| TC         |        |    |
| Statin use (ratio) | 0.198       | 0.181   |

@@

```{r, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "quant90.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "quant90.png"))
```
\section{Results}
The resulting estimate of effects on conditional mean of BMI level may not reflect the size and nature of these effects on lower or upper quantile. For example, in the Figure \@ref(fig:resu1), the conditional mean shows the conditional mean effects of gender on glucose level is about 7. However, the disparity of the gender effects on lower tails is much smaller which is about 3 but the disparity is higher for the upper tail of the distribution somewhat larger than 10 mg/dL.

@@

```{r resu1, out.width = "100%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig.png"))
```
 So,

```{r resu12, out.width = "100%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig11.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig11.png"))
```

```{r resu122, out.width = "80%", fig.align = "center", fig.cap = "Using Cubic splines", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig111.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig111.png"))
```

 therefore
```{r, out.width = "80%", fig.align = "center", fig.cap = "A better figure caption", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig2.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig2.png"))
``` 
So,

 Age effect modeled as a quadratic factor. The age effect  is linear in general, see Figure \@ref(fig:resu3). The slope for the  left quantile is lower than the slope of the right quantile. This can be interpreted as age effect tends to increase over the entire range of conditional distribution of glucose in faster rate for the higher quantile than in the left quantile.In other word, diabetes population have higher rate of change with respect to age if compared to non diabetes population. It was found that there is  a significantly correlation between glucose and age[@tsaousis2014].

@@

```{r resu3, out.width = "80%", fig.align = "center", fig.cap = "Illustration of the quadratic age effect on glucose leveles for four different quantiles of the conditional BMI distribution.  The ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "newfig3.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "newfig3.png"))
```
Using splines 
```{r, out.width = "80%", fig.align = "center", fig.cap = "Age effect modeled using bsplines ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "effect_Age_22.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "effect_Age_22.png"))
```

Marginal quadratic cholesterol effect

@@

Splines modeling

```{r resu33, out.width = "80%", fig.align = "center", fig.cap = " Marginal Cholestrol effect modeled as B splines.  The ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "effect_TC_22.png"))) {
    source(here::here("exploratory-data-analysis_JRT.R"))
}
knitr::include_graphics(here::here("images", "effect_TC_22.png"))
```

Quadratic effect of total cholesterol on the conditional distribution of glucose is convex.  At the lower tail, decreasing glucose level by somewhat around 1.4 up to around 180 Cholesterol level. Then, started to increase glucose levels after TC beyond 180. At higher quantile, the reverse in the cholesterol effects is shifted to higher cholesterol levels which around 200, Figure \@ref(fig:resu4). The relation between total cholesterol and fasting glucose has been studied in [@tsaousis2014],[@chang2011]. They found that there is a positive correlation between the two groups on average.
